---
title: "Loading and initial formatting of Type 2 data"
subtitle: "Data preparation for RBS recovery rate"
author: "Jochen Depestele"
date: today
date-format: "D MMMM YYYY"
format:
  pdf:
    number-sections: true
    tbl-cap-location: top
    include-in-header:
      text: |
        \usepackage{longtable}
execute: 
  warning: false
  echo: false
  message: false
  include: false
editor: visual
--- 

```{r SET-UP R-SESSION}
# These folders are user-specific
data_folder <- "C:/Users/jdepestele/OneDrive - ILVO/000_data/biota/benthos_WKD6SCOPE/2026_WKBENTH4/"
traits_folder <- "C:/Users/jdepestele/OneDrive - ILVO/000_data/biota/traits_benth/"
if(all(dir.exists(paste0(data_folder,c("Type 1","Type2","Type3"))))){
  cat("\n=> Locate data folders (Type 1, Type2, Type3) \nfrom the data call in data_folder !\n ")}
if(!all(dir.exists(c(traits_folder)))){
  cat("\n=> Locate all folders with traits information !\n ")}

in_folder <- "./input"
out_folder <- "./output"

# R packages
library(readxl)
library(data.table)
library(worrms)
library(pbapply)
library(tools)
library(readr)
library(dplyr)
library(kableExtra)

```

```{r PATH OF TYPE 2 DATA}
#| echo: false
## LOAD DATA ----
d_files <- list.files(paste0(data_folder,"Type2"),  pattern = "\\.xlsx$|\\.xls$", full.names = T)
d_files <- d_files[!grepl("_old", d_files)]
d_sheets <- lapply(d_files, function(xfile) {
  sheets <- excel_sheets(xfile)
  data.frame(
    file = basename(xfile),
    has_station_info = "Station information" %in% sheets,
    has_biological_info = "Biological information" %in% sheets
  )
})
d_sheets <- do.call(rbind, d_sheets)
d_sheets[with(d_sheets, !has_station_info | !has_biological_info),] # Which files do NOT have all worksheets?


```

```{r Meta data}
## Meta data ====
meta_lst <- lapply(d_files, function(file)read_excel(file, sheet=excel_sheets(file)[1]))
names(meta_lst) <- gsub("\\.xlsx","",basename(d_files))
meta_dt <- rbindlist(lapply(meta_lst, function(x)x[,c("cl_name","cl_description")]), idcol=T)

```

```{r Station data}
## Stations ====
Stn_lst <- lapply(d_files, function(file)read_excel(file, sheet=excel_sheets(file)[2]))
names(Stn_lst) <- gsub("\\.xlsx","",basename(d_files))

Stn_lst <- lapply(Stn_lst, function(df) {
  names(df) <- tolower(names(df)) # Lowercase column names
  df})
col_freq <- table(unlist(lapply(Stn_lst, names)))
sort(col_freq)
basecols <- names(col_freq[col_freq >= length(Stn_lst)]) # These columns occur in every excel file
# basecols
# mostcols <- sort(names(col_freq[col_freq >= length(Stn_lst)-3]))     # These columns occur in every excel file except 3 files
# mostcols[!mostcols %in% basecols]
basecols <- c(basecols, c("pressure_value",paste0("pressure_value",c("1","2"))), "month")

Stn_dt <- rbindlist(
  lapply(Stn_lst, function(x) x[, intersect(basecols, names(x)), drop = FALSE]), # select basecols when existing
  idcol = T,
  fill = TRUE) # pressure_value1 and pressure_value2 do not exist in most datafiles

to_table1 <- table(Stn_dt$.id, Stn_dt$pressure_type)
to_table2 <- table(Stn_dt$.id, Stn_dt$gear)

Stn_dt[,gear_cat := ifelse(gear %in% unique(Stn_dt$gear)[grepl("trawl",tolower(unique(Stn_dt$gear)))],
                           "trawl",
                           ifelse(gear %in% unique(Stn_dt$gear)[grepl("veen|hamon|day|sm|core",tolower(unique(Stn_dt$gear)))],
                                  "grab",NA))]
c(table(Stn_dt$gear_cat), 
  setNames(sum(table(Stn_dt$gear_cat)), "total"),
  setNames(nrow(Stn_dt), "Nr records Stn_dt")) # All records were sampled with grabs or trawls
Stn_dt[!gear_cat %in% c("trawl","grab")]

selcols <- c(".id","station","replicates","year")
Stn_dt[,..selcols][duplicated(Stn_dt[,..selcols]) | duplicated(Stn_dt[,..selcols], fromLast=T)]
selcols <- c(".id","station","replicates","year","month")
Stn_dt[,..selcols][duplicated(Stn_dt[,..selcols]) | duplicated(Stn_dt[,..selcols], fromLast=T)]

to_table3 <- table(Stn_dt$.id, Stn_dt$gear_cat)

unique(Stn_dt[!is.na(pressure_value1) | !is.na(pressure_value2)]$pressure_type)
Stn_dt[pressure_type == "Sand extraction + fishery"]

# First copy the unchanged columns
DT1 <- copy(Stn_dt[pressure_type == "Sand extraction + fishery"])
DT2 <- copy(Stn_dt[pressure_type == "Sand extraction + fishery"])

# Modify only the needed columns
DT1[, `:=`(
  pressure_type  = "Sand extraction",
  pressure_value = pressure_value1
)]

DT2[, `:=`(
  pressure_type  = "SAR",
  pressure_value = pressure_value2
)]

# nrow(Stn_dt[pressure_type == "Sand extraction + fishery"]) # Split these pressure into two results in a double number of rows
# rbind(DT1, DT2)

Stn_dt <- rbind(Stn_dt[pressure_type != "Sand extraction + fishery"],
                DT1, DT2)

Stn_dt[, c("pressure_value1", "pressure_value2") := NULL]


saveRDS(Stn_dt, paste0(out_folder, "/station_dt.RDS"))

```

# Loading Type 2 data

## Metadata, stations and biological data

### Stations Stn_dt

The stations were loaded from the respective station worksheets and pooled into one datatable. There are four pressure types (@tbl-pressuretypes), which were sampled with a range of grabs or trawls (@tbl-smpl_gears) that were categorized as grab and trawls (@tbl-smpl_gears_revised).

```{r, include=T}
#| label: tbl-pressuretypes
#| tbl-cap: 'Pressure types by input files'

kable(to_table1,
      format = "latex", booktabs = TRUE, escape = T 
      ) %>%
  kable_styling(latex_options = c("hold_position", "striped", "scale_down"))

```

```{r, include=T}
#| label: tbl-smpl_gears
#| tbl-cap: 'Pressure types by input files'

kable(to_table2,
      format = "latex", booktabs = TRUE, escape = T 
      ) %>%
  kable_styling(latex_options = c("hold_position", "striped", "scale_down"))

```

```{r, include=T}
#| label: tbl-smpl_gears_revised
#| tbl-cap: 'Pressure types by input files'

kable(to_table3,
      format = "latex", booktabs = TRUE, escape = T 
      ) %>%
  kable_styling(latex_options = c("hold_position", "striped", "scale_down"))

```

\newpage

```{r Biological data}

B_lst <- lapply(d_files, function(file)read_excel(file, sheet=excel_sheets(file)[3]))
names(B_lst) <- gsub("\\.xlsx","",basename(d_files))
B_lst <- lapply(B_lst, function(df) {
  names(df) <- tolower(names(df)) # Lowercase column names
  df})
col_freq <- table(unlist(lapply(B_lst, names)))
sort(col_freq) # 4 files have taxcode instead of aphiaid

basecols <- names(col_freq[col_freq >= length(Stn_lst)]) # These columns occur in every excel file
basecols <- c(basecols, "aphiaid","taxcode", "month")

B_dt <- rbindlist(
  lapply(B_lst, function(x) x[, intersect(basecols, names(x)), drop = FALSE]), # select basecols when existing
  idcol = T,
  fill = TRUE) # pressure_value1 and pressure_value2 do not exist in most datafiles
B_dt[,.id:=gsub("\\.xlsx","",.id)]

## PRE-PROCESSING: BIOMASS ----
B_dt[,biomass := as.numeric(ifelse(biomass=="NA",NA_character_,biomass))]
datasets_excluded <- unique(B_dt[is.na(biomass)]$.id) # TWO DATASETS HAVE NAs, and ARE FULLY OR PARTIALLY EXCLUDED BECAUSE THEY DO NOT HAVE BIOMASS INFO
# "NS_BfN_beam_btrawling.xlsx" "WMS_IMPEC_btrawling.xlsx"
B_dt <- B_dt[!is.na(biomass)]

## CONVERT BIOMASS VALUES TO THE SAME UNITS ====
biom_unit <- meta_dt[cl_name=="biomass" & .id %in% unique(B_dt$.id)]
biom_unit[, sampled_area_desc := stringr::str_split_i(cl_description, "per ", 2)]
biom_unit[, sampled_area_desc := stringr::str_split_i(sampled_area_desc, "\r" , 1)]
biom_unit[, sampled_area_unit := fifelse(grepl("\\bkm2\\b", sampled_area_desc, TRUE), "km2",
                                         fifelse(grepl("\\bha\\b", sampled_area_desc, TRUE), "ha",
                                                 fifelse(grepl("m²|\\bm2\\b|\\b1m2\\b", sampled_area_desc, TRUE), "m2",
                                                         NA_character_)))]
biom_unit[, sampled_area := fifelse(grepl("^[0-9.]+", sampled_area_desc),
                                    as.numeric(sub("^([0-9.]+).*", "\\1", sampled_area_desc)), 1)]
biom_unit[, sampled_area_sqm := ifelse(grepl("km2",sampled_area_unit),
                                            sampled_area*1e6,ifelse(grepl("ha",sampled_area_unit),
                                                                    sampled_area*10000,ifelse(grepl("m2",sampled_area_unit),
                                                                                              sampled_area,NA_integer_)
                                                                      ))]
selcols <- c(".id","cl_description")
to_table4 <- biom_unit[,..selcols]
B_dt[, biomass_unit := biom_unit$cl_description[match(B_dt$.id, biom_unit$.id, nomatch=NA_integer_)]]
B_dt[, sampled_area_sqm := biom_unit$sampled_area_sqm[match(B_dt$.id, biom_unit$.id, nomatch=NA_integer_)]]
B_dt[,biomass_g_per_sqm := biomass / sampled_area_sqm]
B_dt[,biomass_g_per_sqm := format(biomass_g_per_sqm, scientific = FALSE, trim = TRUE)]


## ASSIGN APHIA-IDs TO ALL TAXA ====
# table(!is.na(B_dt[is.na(aphiaid)]$taxcode)) # When aphiaid is lacking, there's a taxcode
tax_without_id <- sort(unique(B_dt[!is.na(taxcode)]$taxcode))
# tax_aphia_id <- wm_name2id_(name = tax_without_id) # Warnings for some records
rank_order <- c("Kingdom", "Phylum", "Class", "Order", "Infraorder", "Family", "Genus", "Species", "Subspecies", "Variety", "Form")
get_valid_aphia <- function(name) {
  rec <- tryCatch(worrms::wm_records_name(name),
                  error = function(e) NULL,
                  warning = function(w) NULL)
  if (is.null(rec) || length(rec) == 0) return(NA_integer_)
  rec <- rec[rec$status == "accepted" & !is.null(rec$status == "accepted"),]
  if (length(rec) == 0) return(NA_integer_)
  ranks_in_order <- intersect(rank_order, rec$rank)
  if(length(ranks_in_order) == 0) return(rec) # fallback
  return(rec[rec$rank == ranks_in_order[1],])  # Return the AphiaID of the first record with the highest rank
}
tax_aphia_id_lst <- pblapply(tax_without_id, get_valid_aphia)
tax_aphia_id_dt <- rbindlist(lapply(tax_aphia_id_lst, as.data.table), fill = TRUE)
selcols <- c("AphiaID","scientificname","status","rank","valid_name")

tax_without_id[!tax_without_id %in% tax_aphia_id_dt$valid_name] # Missing taxa
wm_records_name("Opisthobranchia")[,c("status","valid_name")] # taxon inquirendum => use higher taxonomic rank, being Class Gastropoda
B_dt[is.na(aphiaid) & taxcode=="Opisthobranchia"]$aphiaid <- wm_records_name("Heterobranchia") %>% filter(status=="accepted") %>% select(AphiaID)
wm_records_name("Pseudocuma")[wm_records_name("Pseudocuma")$rank=="Family",c("valid_name","valid_AphiaID")]  # => add taxon manually
B_dt[is.na(aphiaid) & taxcode=="Pseudocuma"]$aphiaid <- 110384
# tax_aphia_id_dt[valid_name %in% tax_without_id,..selcols]
# tax_aphia_id_dt[!valid_name %in% tax_without_id,..selcols]

B_dt[is.na(aphiaid)]$aphiaid <- tax_aphia_id_dt$AphiaID[match(B_dt[is.na(aphiaid)]$taxcode, tax_aphia_id_dt$valid_name, nomatch=NA_integer_)]

selcols <- c(".id","station_name","replicate","year","month","aphiaid","biomass_g_per_sqm")
B_dt <- B_dt[aphiaid!="NA" & !is.na(aphiaid),..selcols]

saveRDS(B_dt, paste0(out_folder, "/biomass_dt.RDS"))


```

### Biological data B_dt

Biological data are extracted from the third excel sheet from the respective Excel files in the Type 2 data folder. Records which did not included biomass data were excluded, which was the case for all or some records from two datasets: `r datasets_excluded`. Biomass data were converted to the same units using the *cl_description* from the metadata worksheet (@tbl-smpl_area). There were several input Excel files which did not have an Aphia-ID but only a taxon description. The taxon description was used with the `worrms` package to extract an accepted Aphia-ID using the highest to the lowest taxonomic rank. The taxon *Opisthobranchia* was replaced by *Heterobranchia*, and the Family rank of *Pseudocuma* was added manually.

```{r, include=T}
#| label: tbl-smpl_area
#| tbl-cap: 'Sampled area by input Excel file'

kable(to_table4,
      format = "latex", booktabs = TRUE, escape = T 
      ) %>%
  kable_styling(latex_options = c("hold_position", "striped", "scale_down"))

```

\newpage

```{r biological trait databases}
## Load biological trait databases ====
trt_files <- list.files(traits_folder,  pattern = "\\.xlsx$|\\.xls$|\\.csv$", full.names = T, recursive = T)
beauchard_legend <- read_excel(trt_files[6])
trt_files <- trt_files[grepl("Foveau|Vaz|11877|Beauchard_SupplMat\\.xlsx$", trt_files)] # 4 trait databases

read_csv_auto <- function(file) {read_delim(file, delim = NULL, show_col_types = FALSE)}
read_any_file <- function(file) {
  ext <- tolower(file_ext(file))
  switch(ext,
         "csv"  = read_csv_auto(file),
         "xlsx" = read_excel(file),
         "xls"  = read_excel(file),
         stop("Unsupported file type: ", ext))}
trt_sheets <- lapply(trt_files, read_any_file)
names(trt_sheets) <- basename(trt_files)

# Manually select all longevity traits => exclude Foveau
lapply(trt_sheets, names)

```

```{r combined longevity traits}
## COMPILE COMBINED biological trait database ====
# Longevity traits when aphia-id is available (Vaz and Clare):
long <- list()
long[["vaz"]] <- trt_sheets[[2]][,c("aphiaID","meanL1", "meanL1.L3", "meanL3.L10", "meanL10","phylum","class","order","family","genus")]
vaz_NAs <- long[["vaz"]][is.na(long[["vaz"]]$aphiaID),]
vaz_NAs_aphiaids <- rbindlist((pblapply(c("Rissoidae","Hoplostethus","Terebellidae"), get_valid_aphia)))
long[["vaz"]][is.na(long[["vaz"]]$aphiaID),]$aphiaID <- vaz_NAs_aphiaids[status=="accepted"]$AphiaID

long[["clare"]] <- trt_sheets[[3]][,c("AphiaID","l_less_than_1","l1_to_3","l3_to_10","l_more_than_10","Phylum","Class","_Order","Family","Genus")]
long[["clare"]][is.na(long[["clare"]]$AphiaID),]

# Add aphia-id to beauchard's database:
beauchard <- setDT(trt_sheets[[4]][,c("genus","species","LS")])
stan_score <- beauchard_legend[beauchard_legend$trait=="LS",]$standardised_score
beauchard[, longevity := stan_score[LS]]
beauchard <- dcast(beauchard, genus+species~longevity, value.var = "longevity")
beauchard[, sp := paste(genus, species,sep=" ")]
beauchard[, sp := gsub(" sp.","",sp)]
setnames(beauchard, c("0", "0.33", "0.67", "1"),
         c("L1","L3","L10","L10plus"))
beauchard_aphia_id_lst <- pblapply(beauchard$sp, get_valid_aphia)
beauchard_aphia_id_dt <- rbindlist(lapply(beauchard_aphia_id_lst, as.data.table), fill = TRUE)
# names(beauchard_aphia_id_dt)
selcols <- c("valid_name","valid_AphiaID","phylum","class","order","family","genus","scientificname")
beauchard_aphia_id_dt[!is.na(valid_name),..selcols]
missing_taxa <- beauchard$sp[!beauchard$sp %in% beauchard_aphia_id_dt$valid_name] # Missing taxa
# lapply(missing_taxa[!missing_taxa %in% c("Ampeliscanipes","Aporrhais serresianus","Gastrosaccusnifer","Munidaciosa",
#                                          "Pontophilusnosus","Porania pulvillus","Pseudocuma longicorne")], 
#        function(x)wm_records_name(x))
# beauchard$sp[beauchard$sp %in% beauchard_aphia_id_dt$scientificname]

setkey(beauchard, genus, sp)
setkey(beauchard_aphia_id_dt, genus, scientificname)
beauchard <- beauchard_aphia_id_dt[!is.na(valid_name),..selcols][beauchard]
nrow(beauchard[is.na(valid_AphiaID)]) # 26 missing taxa 
beauchard <- beauchard[!is.na(valid_AphiaID)]
selcols <- c("valid_AphiaID","L1","L3","L10","L10plus","phylum","class","order","family","genus","scientificname")
long[["beauchard"]] <- beauchard[,..selcols]
setnafill(long[["beauchard"]],
          type = "const", fill = 0,
          cols = c("L1", "L3", "L10", "L10plus"))

long[["vaz"]]$scientificname <- NA
long[["clare"]]$scientificname <- NA

long <- lapply(long, function(x){
  names(x) <- c("aphiaid","L1","L3","L10","L10plus","phylum","class","order","family","genus","scientificname")
  as.data.table(x)})
long_dt <- rbindlist(long, idcol=T)
long_dt <- long_dt[!is.na(aphiaid)] # redundant

unique_aphias_orig <- length(unique(long_dt$aphiaid))
common_aphiaid <- long_dt[, 
                          uniqueN(.id), by = aphiaid]
overlap_aphias_orig <- table(common_aphiaid$V1) #
# to_plot <- long_dt[aphiaid %in% common_aphiaid[V1 %in% c(2,3), aphiaid]] # To enable comparison of the longevity classes (not done)

round(table(duplicated(long_dt[,-1]))/nrow(long_dt),2)*100 # 10 %  duplicated records
selcols <- c("aphiaid")
round(table(duplicated(long_dt[,..selcols]))/nrow(long_dt[,..selcols]),2)*100 # 35 % duplicated taxa, 
                                                           # implying that trait modalities differed between 65 % of overlapping taxa
long_spp_dt <- long_dt[!is.na(scientificname),
                         lapply(.SD,function(x)mean(x,na.rm=T)), # Take the mean over duplicated taxa
                         by=.(aphiaid, phylum, class, order, family, genus, scientificname),
                         .SDcols=c("L1","L3","L10","L10plus")]
long_genus_dt <- long_dt[,
                         lapply(.SD,function(x)mean(x,na.rm=T)), # Take the mean over duplicated taxa
                         by=.(aphiaid, phylum, class, order, family, genus),
                         .SDcols=c("L1","L3","L10","L10plus")]
long_genus_dt[,scientificname := NA_character_]

# Note that some taxonomic ranks differ even though the aphia-ids are the same ! => they have different longevities
dbls <- long_genus_dt[duplicated(long_genus_dt$aphiaid) | duplicated(long_genus_dt$aphiaid,fromLast = T)]
num_cols <- names(dbls)[sapply(dbls, is.numeric)]
char_cols <- names(dbls)[sapply(dbls, is.character)]
char_cols <- char_cols[!char_cols %in% "aphiaid"]
char_agg_fun <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  x_unique <- unique(x)
  if (length(x_unique) == 1) return(x_unique)
  paste(x_unique, collapse = "%%")
}
dbls_toadd <- dbls[,
                   c(lapply(.SD[, num_cols, with = FALSE], function(x) mean(x, na.rm = TRUE)),
                     lapply(.SD[, char_cols, with = FALSE], char_agg_fun)), 
                   by = .(aphiaid),
                   .SDcols = c(num_cols, char_cols)]
setcolorder(dbls_toadd, names(long_genus_dt))
# Replace duplicates:
long_genus_dt <- rbind(long_genus_dt[!(duplicated(long_genus_dt$aphiaid) | duplicated(long_genus_dt$aphiaid,fromLast = T))],
                       dbls_toadd)

long_dt <- rbind(long_genus_dt, long_spp_dt)

saveRDS(long_dt, paste0(out_folder, "/long_dt.RDS"))

```

```{r adding traits to the B_dt}

selcols <- c(".id","station_name","replicate","year","month","aphiaid","biomass_g_per_sqm")
B_dt <- B_dt[,..selcols]
if(any(duplicated(B_dt))){
  B_dt[duplicated(B_dt) | duplicated(B_dt, fromLast = TRUE)] # See the duplicates
  B_dt <- unique(B_dt) # remove duplicates
}
if(any(duplicated(long_dt))){
  long_dt[duplicated(long_dt) | duplicated(long_dt, fromLast = TRUE)]
  long_dt <- unique(long_dt) # remove duplicates
}

# Add traits using species' scientific names
setkey(B_dt, aphiaid)
setkey(long_dt, aphiaid)

temp_dt <- long_dt[!is.na(scientificname)][B_dt]
dim(B_dt)
dim(temp_dt)
B_spp_dt <- temp_dt[!is.na(scientificname)] # scientific names are given

# Add traits using genus names
B_gen_dt <- temp_dt[is.na(scientificname)]
selcols <- names(B_dt)
B_gen_dt <- B_gen_dt[,..selcols]

setkey(B_gen_dt, aphiaid)
setkey(long_dt, aphiaid)

B_gen_dt <- long_dt[is.na(scientificname)][B_gen_dt]

# Combine datasets with traits, given by spp or genus
B_dt <- rbind(B_spp_dt, B_gen_dt)

setcolorder(B_dt, c(".id","station_name","replicate","year","month","aphiaid","biomass_g_per_sqm","L1","L3","L10","L10plus","phylum","class","order","family","genus","scientificname"))
saveRDS(B_dt, paste0(out_folder, "/biomass_long_dt.RDS"))


```

## Biological traits

Four biological traits databases were loaded: Foveau et al (2020), Beauchard et al (2022), Clare et al (2022) and Vaz et al (2022.) Foveau et al (2020) did not include longevity as a trait. Three missing Aphia-IDs were added to Vaz et al (2022), for the Families Rissoidae and Terebellidae, and for the Genus Hoplostethus. Aphia-IDs were added to Beauchard et al (2022). All three databases had the four same trait modalities for longevity: longevity $\leq 1$, 1-3, 3-10 and $\geq 10$. All data were pooled into one datatable, and records without Aphia-ID were excluded. Trait scores were averaged over the records at species level (which is in principle redundant as it only comprises the database from Beauchard et al 2022), and then at genus level. For some records the Genera matched, but the lower taxonomic rank did not, e.g. the genus Calliactis was either categorized as the class Anthozoa, or Octocorallia. Both names were merged, and the duplicated records were replaced using the merged taxonomic name. Biological trait modalities from the combined longevity database were added to the biological biomass data.
