---
title: "Thresholds metanalysis"
author: "Aurélien Boyé - Ifremer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmdformats::readthedown:
    lightbox: TRUE
    gallery: TRUE
    thumbnails: FALSE # if TRUE display content images as thumbnails
    toc_depth: 3
    use_bookdown: TRUE
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Packages

```{r, message = FALSE}
library(tidyverse)
library(readxl)
library(glmm.hp)
library(betareg)
library(ggridges)
library(broom)
library(mgcv)
library(gam.hp)
library(upset.hp)
library(DHARMa)
library(ggpointdensity)
library(viridis)
```

# Using type 3 data

```{r}

# Load the data
type3_thresholds <- read_excel("06. Data/Data_type_3_threholds_pressure_state.xlsx", sheet = 1)

# Classify infaunal datasets
infauna_datasets <- c("NS_BEhabitats", "NS_NLhabitats", "NS_DKhabitats", "CS_NS_UKhabitats")
 
# Regions (dataset lists provided)
wms_datasets <- c("WMS_IEOESPhabitats", "WMS_EShabitats", "WMS_FRMEDITS","WMS_ISCMS_IRBIMCNR", "WMS_NOURMED", "WMS_APPEALMED")
bobic_cs_datasets <- c("BoBIC_IberianChabitats", "BoBIC_GulfofCadizhabitats", "BoBIC_CGFS","FR_ORHAGO", "CS_EVHOE")
gns_datasets <- c("CS_NS_IBTSFR","NS_BEhabitats", "NS_NLhabitats", "NS_DKhabitats", "CS_NS_UKhabitats")
 
# Extract and format state thresholds
type3_state_thresholds <- type3_thresholds %>%
  select(-DC_P.TH, -NV_P.TH, -D2D_P.TH) %>%
  gather(THRESHOLD_METHOD, value, -DATASET, -INDICATOR, -BHT, -N, -PVALUE) %>%
  mutate(REGION = case_when(
      DATASET %in% wms_datasets      ~ "Western Mediterranean Sea",
      DATASET %in% bobic_cs_datasets ~ "Bay of Biscay & Iberian Coast + Celtic Seas",
      DATASET %in% gns_datasets      ~ "Greater North Sea",
      TRUE                           ~ NA_character_
    ),
    FAUNA = if_else(DATASET %in% infauna_datasets, "Infauna", "Epifauna")
  ) %>%
  mutate(value = as.numeric(value)) %>%
  mutate_if(is.character,as.factor)

# Extract and format pressure thresholds
type3_pressure_thresholds <- type3_thresholds %>%
  select(-DC_S.TH, -NV_S.TH, -D2D_S.TH) %>%
  gather(THRESHOLD_METHOD, value, -DATASET, -INDICATOR, -BHT, -N, -PVALUE) %>%
  mutate(
    REGION = case_when(
      DATASET %in% wms_datasets      ~ "Western Mediterranean Sea",
      DATASET %in% bobic_cs_datasets ~ "Bay of Biscay & Iberian Coast + Celtic Seas",
      DATASET %in% gns_datasets      ~ "Greater North Sea",
      TRUE                           ~ NA_character_
    ),
    FAUNA = if_else(DATASET %in% infauna_datasets, "Infauna", "Epifauna")
  ) %>%
  mutate(value = as.numeric(value)) %>%
  mutate_if(is.character,as.factor)
```

## Distribution of state thresholds

```{r}

ggplot(aes(x = value, y = DATASET), data = type3_state_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )

ggplot(aes(x = value, y = INDICATOR), data = type3_state_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )


ggplot(aes(x = value, y = BHT), data = type3_state_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )

ggplot(aes(x = value, y = THRESHOLD_METHOD), data = type3_state_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )

ggplot(aes(x = value, y = REGION), data = type3_state_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )

ggplot(aes(x = value, y = FAUNA), data = type3_state_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )

```

## Distribution of pressure thresholds

```{r}

ggplot(aes(x = value, y = DATASET), data = type3_pressure_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )

ggplot(aes(x = value, y = INDICATOR), data = type3_pressure_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )


ggplot(aes(x = value, y = BHT), data = type3_pressure_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )

ggplot(aes(x = value, y = THRESHOLD_METHOD), data = type3_pressure_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )

ggplot(aes(x = value, y = REGION), data = type3_pressure_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )

ggplot(aes(x = value, y = FAUNA), data = type3_pressure_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )

```

## Variance partitioning of pressure thresholds

```{r}
detect_glm_family(type3_pressure_thresholds$value)
```

Best model is a gamma one, but variance partitioning not available so I first compare with a poisson regression

## With dataset as predictor

```{r}
thresholds_model_gamma <- glm(value ~ DATASET + FAUNA + REGION + INDICATOR + BHT + N + THRESHOLD_METHOD, family = Gamma(link = "log") , data = type3_pressure_thresholds)

summary(thresholds_model_gamma)

model_residuals <- simulateResiduals(fittedModel = thresholds_model_gamma)
plot(model_residuals)

coef_gamma <- tidy(thresholds_model_gamma, conf.int = TRUE) %>%
  mutate(signif = if_else(p.value < 0.05, TRUE, FALSE))

coef_gamma

ggplot(mapping = aes(x = estimate, y = term, fill = p.value, shape = signif), data = coef_gamma) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_point(size = 3)
```


```{r}
thresholds_model_poisson <- glm(value ~ DATASET + FAUNA + REGION + INDICATOR + BHT + N + THRESHOLD_METHOD, family = poisson(link = "log") , data = type3_pressure_thresholds)

summary(thresholds_model_poisson)

model_residuals <- simulateResiduals(fittedModel = thresholds_model_poisson)
plot(model_residuals)

# Extract the coef
coef_poisson <- tidy(thresholds_model_poisson, conf.int = TRUE) %>%
  mutate(signif = if_else(p.value < 0.05, TRUE, FALSE))

coef_poisson

ggplot(mapping = aes(x = estimate, y = term, fill = p.value, shape = signif), data = coef_poisson) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_point(size = 3)
```

```{r}
plot(predict(thresholds_model_poisson, type = "response") ~ predict(thresholds_model_gamma, type = "response"))
abline(a = 0, b = 1)

plot(coef_poisson$estimate ~ coef_gamma$estimate)
abline(a = 0, b = 1)
```

Seems ok, qualitatively results are similar, although Gamma model behave better. I still go on with variance partitioning for the poisson model given similar coefficients and predictions.

```{r}
variance_partition <- glmm.hp(thresholds_model_poisson, iv = NULL, type = "adjR2", commonality = FALSE)

variance_partition

plot(variance_partition)

variance_partition_common <- glmm.hp(thresholds_model_poisson, iv = NULL, type = "adjR2", commonality = TRUE)

vp <- variance_partition_common$delta
hp <- variance_partition$delta
upset.hp(vp, hp, col = "cvd", digits = 1, nVar = 19)
```

### If I remove dataset as predictor

```{r}
thresholds_model_poisson <- glm(value ~ FAUNA + REGION + INDICATOR + BHT + N + THRESHOLD_METHOD, family = poisson(link = "log") , data = type3_pressure_thresholds)

summary(thresholds_model_poisson)

model_residuals <- simulateResiduals(fittedModel = thresholds_model_poisson)
plot(model_residuals)

# Extract the coef
coef_poisson <- tidy(thresholds_model_poisson, conf.int = TRUE) %>%
  mutate(signif = if_else(p.value < 0.05, TRUE, FALSE))

coef_poisson

ggplot(mapping = aes(x = estimate, y = term, fill = p.value, shape = signif), data = coef_poisson) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_point(size = 3)

variance_partition <- glmm.hp(thresholds_model_poisson, iv = NULL, type = "adjR2", commonality = FALSE)

variance_partition

plot(variance_partition)

variance_partition_common <- glmm.hp(thresholds_model_poisson, iv = NULL, type = "adjR2", commonality = TRUE)

vp <- variance_partition_common$delta
hp <- variance_partition$delta
upset.hp(vp, hp, col = "cvd", digits = 1)
```

## Variance partitioning of state thresholds

```{r}
detect_glm_family(type3_state_thresholds$value)
```

Best model is a beta regression

> Note here that I cannot put both DATASET and REGION+INFAUNA in the beta model (except with the gam), due to convergence issue. Hence I test them alternatively in the model then use them both in the gam in the end. 

### With dataset as predictor

```{r}

thresholds_model_beta <- betareg(value ~ DATASET + INDICATOR + BHT + N + THRESHOLD_METHOD, data = type3_state_thresholds)
summary(thresholds_model_beta)

par(mfrow=(c(2,2)))
plot(thresholds_model_beta)
par()

coef_beta <- tidy(thresholds_model_beta, conf.int = TRUE) %>%
  mutate(signif = if_else(p.value < 0.05, TRUE, FALSE))

coef_beta

ggplot(mapping = aes(x = estimate, y = term, fill = p.value, shape = signif), data = coef_beta %>% filter(component == "mean")) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_point(size = 3)

```

```{r}
thresholds_model_beta_gam_lm <- gam(
  value ~ DATASET + INDICATOR + BHT + N + THRESHOLD_METHOD,
  family = betar(link = "logit"),
  data = type3_state_thresholds,
  method = "REML"
)

thresholds_model_beta_gam_lm

coef_beta <- summary(thresholds_model_beta_gam_lm)$p.table %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    statistic = `z value`,
    p.value = `Pr(>|z|)`
  ) %>%
  mutate(
    conf.low  = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    signif = p.value < 0.05
  )

ggplot(mapping = aes(x = estimate, y = term, fill = p.value, shape = signif), data = coef_beta) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_point(size = 3)

variance_partition_beta <- gam.hp(
  thresholds_model_beta_gam_lm,
  iv = NULL,
  type = "dev",          # IMPORTANT
  commonality = FALSE
)

variance_partition_beta
plot(variance_partition_beta)

variance_partition_beta_common <- gam.hp(
  thresholds_model_beta_gam_lm,
  iv = NULL,
  type = "dev",          # IMPORTANT
  commonality = TRUE
)

vp <- variance_partition_beta_common$commonality.analysis
hp <- variance_partition_beta$hierarchical.partitioning
upset.hp(vp, hp, col = "cvd", digits = 1, nVar = 9)
```

### With region and fauna (removing dataset)

```{r}

thresholds_model_beta <- betareg(value ~ REGION + FAUNA + INDICATOR + BHT + N + THRESHOLD_METHOD, data = type3_state_thresholds)
summary(thresholds_model_beta)

plot(thresholds_model_beta)

coef_beta <- tidy(thresholds_model_beta, conf.int = TRUE) %>%
  mutate(signif = if_else(p.value < 0.05, TRUE, FALSE))

coef_beta

ggplot(mapping = aes(x = estimate, y = term, fill = p.value, shape = signif), data = coef_beta %>% filter(component == "mean")) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_point(size = 3)

```

```{r}
thresholds_model_beta_gam_lm <- gam(
  value ~  REGION + FAUNA + INDICATOR + BHT + N + THRESHOLD_METHOD,
  family = betar(link = "logit"),
  data = type3_state_thresholds,
  method = "REML"
)

thresholds_model_beta_gam_lm

coef_beta <- summary(thresholds_model_beta_gam_lm)$p.table %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    statistic = `z value`,
    p.value = `Pr(>|z|)`
  ) %>%
  mutate(
    conf.low  = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    signif = p.value < 0.05
  )

ggplot(mapping = aes(x = estimate, y = term, fill = p.value, shape = signif), data = coef_beta) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_point(size = 3)

variance_partition_beta <- gam.hp(
  thresholds_model_beta_gam_lm,
  iv = NULL,
  type = "dev",          # IMPORTANT
  commonality = FALSE
)

variance_partition_beta
plot(variance_partition_beta)

variance_partition_beta_common <- gam.hp(
  thresholds_model_beta_gam_lm,
  iv = NULL,
  type = "dev",          # IMPORTANT
  commonality = TRUE
)

vp <- variance_partition_beta_common$commonality.analysis
hp <- variance_partition_beta$hierarchical.partitioning
upset.hp(vp, hp, col = "cvd", digits = 1)
```

### With both

```{r}
thresholds_model_beta_gam_lm <- gam(
  value ~  DATASET + REGION + FAUNA + INDICATOR + BHT + N + THRESHOLD_METHOD,
  family = betar(link = "logit"),
  data = type3_state_thresholds,
  method = "REML"
)

thresholds_model_beta_gam_lm

coef_beta <- summary(thresholds_model_beta_gam_lm)$p.table %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    statistic = `z value`,
    p.value = `Pr(>|z|)`
  ) %>%
  mutate(
    conf.low  = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    signif = p.value < 0.05
  )

ggplot(mapping = aes(x = estimate, y = term, fill = p.value, shape = signif), data = coef_beta) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_point(size = 3)

variance_partition_beta <- gam.hp(
  thresholds_model_beta_gam_lm,
  iv = NULL,
  type = "dev",          # IMPORTANT
  commonality = FALSE
)

variance_partition_beta
plot(variance_partition_beta)

variance_partition_beta_common <- gam.hp(
  thresholds_model_beta_gam_lm,
  iv = NULL,
  type = "dev",          # IMPORTANT
  commonality = TRUE
)

vp <- variance_partition_beta_common$commonality.analysis
hp <- variance_partition_beta$hierarchical.partitioning
upset.hp(vp, hp, col = "cvd", digits = 1, nVar = 16)
```

---

# Using type 1 data

```{r}
# Load the data
type1_thresholds <- read.csv("06. Data/aurelien_data.csv")

for (i in c("station","gear","indicator","data_type","substrate","method")){
  p <- ggplot(aes(x = threshold, y = .data[[i]]), data = type1_thresholds) +
  geom_density_ridges(
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.7,
  )

  plot(p)
}

ggplot(aes(x = depth, y = threshold), data = type1_thresholds) +
  geom_pointdensity() +
  scale_color_viridis()

```

```{r}
detect_glm_family(type1_thresholds$threshold)

thresholds_model_beta_gam_lm <- gam(
  threshold ~  station + gear + indicator + data_type + MSFD_BBHT + depth + method,
  family = betar(link = "logit"),
  data = type1_thresholds,
  method = "REML"
)

thresholds_model_beta_gam_lm

coef_beta <- summary(thresholds_model_beta_gam_lm)$p.table %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    statistic = `z value`,
    p.value = `Pr(>|z|)`
  ) %>%
  mutate(
    conf.low  = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    signif = p.value < 0.05
  )

ggplot(mapping = aes(x = estimate, y = term, fill = p.value, shape = signif), data = coef_beta) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_point(size = 3)

variance_partition_beta <- gam.hp(
  thresholds_model_beta_gam_lm,
  iv = NULL,
  type = "dev",          # IMPORTANT
  commonality = FALSE
)

variance_partition_beta
plot(variance_partition_beta)

variance_partition_beta_common <- gam.hp(
  thresholds_model_beta_gam_lm,
  iv = NULL,
  type = "dev",          # IMPORTANT
  commonality = TRUE
)

vp <- variance_partition_beta_common$commonality.analysis
hp <- variance_partition_beta$hierarchical.partitioning
upset.hp(vp, hp, col = "cvd", digits = 1, nVar = 7)
```


---

# Session info

```{r}
sessionInfo()
```

